A Comparison of R Tools for Nonlinear Least Squares Modeling

### 2023-10-18 JNABRev20231018/ComparingNLStools.tex
2nd revision

\hypertarget{the-nls-function}{%
\subsection{\texorpdfstring{The \texttt{nls()} function}{The nls() function}}\label{the-nls-function}}

\texttt{nls()} is the tool for estimating nonlinear statistical models in base R, the
primary software distribution
from the Comprehensive R Archive Network (\emph{\url{https://cran.r-project.org}}).
The function dates to the 1980s and the work related to Bates and Watts (1988) in S (see
\emph{\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}}).

@@

\hypertarget{scope-of-our-comparison}{%
\subsection{Scope of our comparison}\label{scope-of-our-comparison}}

@@

\hypertarget{principal-messages}{%
\subsubsection{Principal messages}\label{principal-messages}}

@@

\hypertarget{some-other-cran-packages-for-nonlinear-modeling}{%
\subsection{Some other CRAN packages for nonlinear modeling}\label{some-other-cran-packages-for-nonlinear-modeling}}

@@

\texttt{NISTnls} ((\emph{NISTnls: Nonlinear Least Squares Examples from Nist} 2012)) -- This package provides R code and data for a set of
(numerically ill-conditioned) nonlinear least squares problems from the U.S. National
Institute for Standards and Technology. These may not represent real-world situations.

@@

\texttt{optimx} (Nash and Varadhan (2011)) -- This provides
optimizers that can be applied to minimize a nonlinear function which could be
a nonlinear sum of
squares. NOT generally recommended if nonlinear
least squares programs can be easily used, but provides a check and alternative
solvers.

\hypertarget{an-illustrative-example}{%
\subsection{An illustrative example}\label{an-illustrative-example}}

The Hobbs weed infestation problem (Nash (1979, 120)) is a growth curve modeling
task that seems straightforward but is quite nasty. Its very succinct statement
provides the ``short reproducible example'' much requested on R mailing lists.
This is a real problem from a field researcher. The data and graph follow.

@@

@@

The usual suggestion is that \(D\) be replaced by a multiple of the unit matrix or else
a multiple of the diagonal part of \(J' J\). In low precision, some elements of
\(J' J\) could underflow to zero (Nash (1977)), and a linear combination of both
choices is an effective compromise. Various choices for
\(D\), as well as a possible line search along the
direction \(\delta\) rather than a unit step (Hartley (1961)),
give rise to several variant
algorithms. ``Marquardt's method'' is a family of methods. Fortunately, most choices
work well.

\hypertarget{problem-setup}{%
\subsubsection{Problem setup}\label{problem-setup}}

@@

It is also possible to provide R functions for the residual and Jacobian.
This is usually more work for the user if the formula setup is possible.
To illustrate, we show the functions for the unscaled 3 parameter logistic.

\begin{verbatim}
# Logistic3U
hobbs.res  <-  function(x){ # scaled Hobbs weeds problem -- residual
  # This variant uses looping
  if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
  y  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
           38.558, 50.156, 62.948, 75.995, 91.972)
  tt  <-  1:12
  res  <-  x[1]/(1+x[2]*exp(-x[3]*tt)) - y
}

hobbs.jac  <-  function(x) { # scaled Hobbs weeds problem -- Jacobian
  jj  <-  matrix(0.0, 12, 3)
  tt  <-  1:12
  yy  <-  exp(-x[3]*tt)
  zz  <-  1.0/(1+x[2]*yy)
  jj[tt,1]   <-   zz
  jj[tt,2]   <-   -x[1]*zz*zz*yy
  jj[tt,3]   <-   x[1]*zz*zz*yy*x[2]*tt
  attr(jj, "gradient") <- jj
  jj
}
\end{verbatim}

\hypertarget{estimation-of-models-specified-as-formulas}{%
\subsection{Estimation of models specified as formulas}\label{estimation-of-models-specified-as-formulas}}

@@

Here we see the infamous ``singular gradient'' termination message of \texttt{nls()}.

\hypertarget{solution-attempts-with-nlsr}{%
\subsubsection{Solution attempts with nlsr}\label{solution-attempts-with-nlsr}}

@@

\hypertarget{solution-attempts-with-minpack.lm}{%
\subsubsection{Solution attempts with minpack.lm}\label{solution-attempts-with-minpack.lm}}

@@

\hypertarget{solution-attempts-with-gslnls}{%
\subsubsection{Solution attempts with gslnls}\label{solution-attempts-with-gslnls}}

@@

\hypertarget{comparison-notes-for-formula-setup-solutions}{%
\subsubsection{Comparison notes for formula-setup solutions}\label{comparison-notes-for-formula-setup-solutions}}

@@

\hypertarget{functional-specification-of-problems}{%
\subsection{Functional specification of problems}\label{functional-specification-of-problems}}

@@

\hypertarget{design-goals-termination-tests-and-output-objects}{%
\subsection{Design goals, termination tests, and output objects}\label{design-goals-termination-tests-and-output-objects}}

@@

\hypertarget{returned-results-of-nls-and-other-tools}{%
\subsubsection{Returned results of nls() and other tools}\label{returned-results-of-nls-and-other-tools}}

@@

\hypertarget{when-to-compute-ancillary-information}{%
\subsubsection{When to compute ancillary information}\label{when-to-compute-ancillary-information}}

Tools that produce a class \texttt{nls} output object create a rich set of functions
and structures that are then used in a variety of modeling tasks, including the
least squares solution. By contrast,
\texttt{nlsr} computes quantities as they are requested or needed, with additional
features in separate functions. For example, the singular values of the Jacobian
are actually computed in the \texttt{print} and \texttt{summary} methods for the result.
These two approaches
lead to different consequences for performance and how features are
provided. \texttt{nlsr} has antecedents in the methods of Nash (1979), where storage for
data and programs was at a ridiculous premium in the small computers of the era.
Thus the code in \texttt{nlsr} is likely of value for workers to copy and modify
for customized tools.

\hypertarget{jacobian-calculation}{%
\subsection{Jacobian calculation}\label{jacobian-calculation}}

@@

\hypertarget{jacobian-code-in-selfstart-models}{%
\subsubsection{Jacobian code in selfStart models}\label{jacobian-code-in-selfstart-models}}

@@

We are also surprised that the
analytic expressions for the Jacobian (``gradient'') in the \texttt{SSLogis} function
and others save quantities in ``hidden''
variables, i.e., with names preceded by ``.''.
These are then not displayed by the \texttt{ls()}
command, making them difficult to access by users who may wish to create
their own selfStart model via copy and edit.
Interactive tools, such as ``visual fitting'' (Nash and Velleman (1996)) might be
worth considering as another way to find starting parameters, but we know of no
R capability of this type.

@@

\hypertarget{bounds-constraints-on-parameters}{%
\subsection{Bounds constraints on parameters}\label{bounds-constraints-on-parameters}}

@@

\hypertarget{philosophical-considerations}{%
\subsubsection{Philosophical considerations}\label{philosophical-considerations}}

@@

\hypertarget{fixed-parameters-masks}{%
\subsubsection{Fixed parameters (masks)}\label{fixed-parameters-masks}}

@@

\hypertarget{stabilization-of-gauss-newton-computations}{%
\subsection{Stabilization of Gauss-Newton computations}\label{stabilization-of-gauss-newton-computations}}

@@

\hypertarget{programming-language}{%
\subsubsection{Programming language}\label{programming-language}}

@@

\hypertarget{data-sources-for-problems}{%
\subsection{Data sources for problems}\label{data-sources-for-problems}}

@@

\hypertarget{feature-subsetting}{%
\subsubsection{Feature: Subsetting}\label{feature-subsetting}}

@@

\hypertarget{feature-na.action-missing-value-treatment}{%
\subsubsection{Feature: na.action (missing value treatment)}\label{feature-na.action-missing-value-treatment}}

@@

\hypertarget{feature-model-frame}{%
\subsection{Feature: model frame}\label{feature-model-frame}}

@@

\hypertarget{weights-on-observations}{%
\subsection{Weights on observations}\label{weights-on-observations}}

@@

\hypertarget{weights-in-returned-functions-from-nls}{%
\subsubsection{Weights in returned functions from nls()}\label{weights-in-returned-functions-from-nls}}

@@

\hypertarget{minor-issues-with-nonlinear-least-squares-tools}{%
\subsection{Minor issues with nonlinear least-squares tools}\label{minor-issues-with-nonlinear-least-squares-tools}}

\hypertarget{interim-output-from-the-port-algorithm}{%
\subsubsection{Interim output from the ``port'' algorithm}\label{interim-output-from-the-port-algorithm}}

@@

\hypertarget{failure-to-return-the-best-result-achieved}{%
\subsubsection{Failure to return the best result achieved}\label{failure-to-return-the-best-result-achieved}}

@@

\hypertarget{estimating-models-that-are-partially-linear}{%
\subsection{Estimating models that are partially linear}\label{estimating-models-that-are-partially-linear}}

@@

\hypertarget{models-with-indexed-parameters}{%
\subsection{Models with indexed parameters}\label{models-with-indexed-parameters}}

Some models have several common parameters and
others that are tied to particular cases.
The \textbf{man} file for \texttt{nls()} includes an example of a situation in which
parameters are indexed but which uses the ``plinear'' option as an added complication.
Running this example reveals that the answers for the parameters are NOT indexed
as in a vector.
That is, we do not see \texttt{a{[}1{]},\ a{[}2{]},\ a{[}3{]}} but \texttt{a1,\ a2,\ a3}.
This is no doubt because programming for indexed parameters is challenging.
We note that there are capabilities in packages for mixed-effects modeling
such as \texttt{nlme} (Pinheiro et al. (2013)), \texttt{bbmle} (Bolker and Team (2013)), and \texttt{lme4} (Bates et al. (2015)) for
estimating models of such type.

\hypertarget{tests-and-use-case-examples}{%
\subsection{Tests and use-case examples}\label{tests-and-use-case-examples}}

@@

\hypertarget{documentation-and-resources}{%
\subsection{Documentation and resources}\label{documentation-and-resources}}

@@

\hypertarget{future-of-nonlinear-model-estimation-in-r}{%
\subsection{Future of nonlinear model estimation in R}\label{future-of-nonlinear-model-estimation-in-r}}

@@

\hypertarget{acknowledgments}{%
\section{Acknowledgments}\label{acknowledgments}}

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-R-onls-manual}{}%
Andrej-Nikolai Spiess. 2022. \emph{Onls: Orthogonal Nonlinear Least-Squares Regression}. \url{https://CRAN.R-project.org/package=onls}.

\leavevmode\hypertarget{ref-bateswatts}{}%
Bates, D. M., and D. G. Watts. 1988. \emph{Nonlinear Regression Analysis and Its Applications}. Wiley.

\leavevmode\hypertarget{ref-p-lme4}{}%
Bates, Douglas, Martin Mächler, Ben Bolker, and Steve Walker. 2015. ``Fitting Linear Mixed-Effects Models Using lme4.'' \emph{Journal of Statistical Software} 67 (1): 1--48. \url{https://doi.org/10.18637/jss.v067.i01}.

\leavevmode\hypertarget{ref-BatesWatts81}{}%
Bates, Douglas M., and Watts, Donald G. 1981. ``A Relative Offset Orthogonality Convergence Criterion for Nonlinear Least Squares.'' \emph{Technometrics} 23 (2): 179--83.

\leavevmode\hypertarget{ref-p-nlstools}{}%
Baty, Florent, and Marie-Laure Delignette-Muller. 2013. \emph{Nlstools: Tools for Nonlinear Regression Diagnostics}.

\leavevmode\hypertarget{ref-nlsMicrobio-manual}{}%
---------. 2014. \emph{NlsMicrobio: Data Sets and Nonlinear Regression Models Dedicated to Predictive Microbiology}.

\leavevmode\hypertarget{ref-p-bbmle}{}%
Bolker, Ben, and R Development Core Team. 2013. \emph{Bbmle: Tools for General Maximum Likelihood Estimation}. \url{http://CRAN.R-project.org/package=bbmle}.

\leavevmode\hypertarget{ref-gslnls23}{}%
Chau, Joris. 2023. \emph{Gslnls: GSL Nonlinear Least-Squares Fitting}. \url{https://CRAN.R-project.org/package=gslnls}.

\leavevmode\hypertarget{ref-nlshelper-manual}{}%
Duursma, Remko. 2017. \emph{Nlshelper: Convenient Functions for Non-Linear Regression}. \url{https://CRAN.R-project.org/package=nlshelper}.

\leavevmode\hypertarget{ref-minpacklm12}{}%
Elzhov, Timur V., Katharine M. Mullen, Andrej-Nikolai Spiess, and Ben Bolker. 2012. \emph{Minpack.lm: R Interface to the Levenberg-Marquardt Nonlinear Least-Squares Algorithm Found in Minpack, Plus Support for Bounds}. R Project for Statistical Computing. \url{http://CRAN.R-project.org/package=minpack.lm}.

\leavevmode\hypertarget{ref-GSL-manual}{}%
Galassi, Mark, Jim Davies, James Theiler, Brian Gough, and Gerard Jungman. 2009. \emph{GNU Scientific Library - Reference Manual, Third Edition, for Gsl Version 1.12 (3. Ed.).}

\leavevmode\hypertarget{ref-Bioconductor}{}%
Gentleman, Robert C., Vincent J. Carey, Douglas M. Bates, Ben Bolstad, Marcel Dettling, Sandrine Dudoit, Byron Ellis, et al. 2004. ``Bioconductor: Open Software Development for Computational Biology and Bioinformatics.'' \emph{Genome Biology} 5 (R80). \url{https://doi.org/10.1186/gb-2004-5-10-r80}.

\leavevmode\hypertarget{ref-Golub1973}{}%
Golub, G. H., and V. Pereyra. 1973. ``The Differentiation of Pseudo-Inverses and Nonlinear Least Squares Problems Whose Variables Separate.'' \emph{SIAM Journal of Numerical Analysis} 10 (2): 413--32.

\leavevmode\hypertarget{ref-nls2-22}{}%
Grothendieck, G. 2022. \emph{Nls2: Non-Linear Regression with Brute Force}. \url{https://CRAN.R-project.org/package=nls2}.

\leavevmode\hypertarget{ref-Hartley1961}{}%
Hartley, H. O. 1961. ``The Modified Gauss-Newton Method for the Fitting of Non-Linear Regression Functions by Least Squares.'' \emph{Technometrics} 3: 269--80.

\leavevmode\hypertarget{ref-Huet2004}{}%
Huet, S., A. Bouvier, M.-A. Poursat, and E. Jolivet. 2004. \emph{Statistical Tools for Nonlinear Regression: A Practical Guide with S-PLUS Examples, 2nd Edition}. Berlin \& New York: Springer-Verlag.

\leavevmode\hypertarget{ref-nlsr2023manual}{}%
John C Nash, and Duncan Murdoch. 2023. \emph{nlsr: Functions for Nonlinear Least Squares Solutions}.

\leavevmode\hypertarget{ref-tvrdik2007}{}%
Josef Tvrdík and Ivan Křivý and Ladislav Mišík. 2007. ``Adaptive Population-Based Search: Application to Estimation of Nonlinear Regression Parameters.'' \emph{Computational Statistics \& Data Analysis} 52 (2): 713--24. \url{https://doi.org/10.1016/j.csda.2006.10.014}.

\leavevmode\hypertarget{ref-Levenberg1944}{}%
Levenberg, Kenneth. 1944. ``A Method for the Solution of Certain Non-Linear Problems in Least Squares.'' \emph{Quarterly of Applied Mathematics} 2: 164--68.

\leavevmode\hypertarget{ref-Marquardt1963}{}%
Marquardt, Donald W. 1963. ``An Algorithm for Least-Squares Estimation of Nonlinear Parameters.'' \emph{SIAM Journal on Applied Mathematics} 11 (2): 431--41.

\leavevmode\hypertarget{ref-MiguezNLRAA2021}{}%
Miguez, Fernando. 2021. \emph{nlraa: Nonlinear Regression for Agricultural Applications}. \url{https://CRAN.R-project.org/package=nlraa}.

\leavevmode\hypertarget{ref-jn77ima}{}%
Nash, John C. 1977. ``Minimizing a Nonlinear Sum of Squares Function on a Small Computer.'' \emph{Journal of the Institute for Mathematics and Its Applications} 19: 231--37.

\leavevmode\hypertarget{ref-cnm79}{}%
---------. 1979. \emph{Compact Numerical Methods for Computers: Linear Algebra and Function Minimisation}. Bristol: Adam Hilger.

\leavevmode\hypertarget{ref-NashWires22}{}%
---------. 2022. ``Function minimization and nonlinear least squares in R.'' \emph{WIREs Computational Statistics}, no. e1580. \url{https://doi.org/https://doi.org/10.1002/wics.1580}.

\leavevmode\hypertarget{ref-PkgFromRbase22}{}%
Nash, John C., and Arkajyoti Bhattacharjee. 2022. ``Making a Package from Base R Files.'' \emph{R-Bloggers}, January. \url{https://www.r-bloggers.com/2022/01/making-a-package-from-base-r-files/}.

\leavevmode\hypertarget{ref-p-optimx}{}%
Nash, John C, and Ravi Varadhan. 2011. \emph{\textup{Optimx}: A Replacement and Extension of the optim() Function}. Nash Information Services Inc.; Johns Hopkins University.

\leavevmode\hypertarget{ref-nash1996nonlinear}{}%
Nash, John C., and Paul Velleman. 1996. ``Nonlinear Estimation Combining Visual Fitting with Optimization Methods.'' In \emph{Proceedings of the Section on Physical and Engineering Sciences of the American Statistical Association}, 256--61. American Statistical Association.

\leavevmode\hypertarget{ref-NISTnls-manual}{}%
\emph{NISTnls: Nonlinear Least Squares Examples from Nist}. 2012. \url{https://CRAN.R-project.org/package=NISTnls}.

\leavevmode\hypertarget{ref-OlearyRust13}{}%
O'Leary, Dianne P., and Bert W. Rust. 2013. ``Variable Projection for Nonlinear Least Squares Problems.'' \emph{Computational Optimization and Applications} 54 (3): 579--93.

\leavevmode\hypertarget{ref-nlsmultstart}{}%
Padfield, Daniel, and Granville Matheson. 2020. \emph{Nls.multstart: Robust Non-Linear Regression Using Aic Scores}. \url{https://CRAN.R-project.org/package=nls.multstart}.

>>>>

\subsection{\texorpdfstring{The \texttt{nls()} function}{The nls() function}}\label{the-nls-function}

\texttt{nls()} is the tool for estimating nonlinear statistical models in base R, the
primary software distribution
from the Comprehensive R Archive Network (\emph{\url{https://cran.r-project.org}}).
The function dates to the 1980s and the work related to D. M. Bates and Watts (1988) in S (see
\emph{\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}}).

@@

\subsection{Scope of our comparison}\label{scope-of-our-comparison}

@@

\subsubsection{Principal messages}\label{principal-messages}

@@

\subsection{Some other CRAN packages for nonlinear modeling}\label{some-other-cran-packages-for-nonlinear-modeling}

@@

\texttt{NISTnls} (National Institutes for Standards and R port by Douglas Bates (2012)) -- This package provides R code and data for a set of
(numerically ill-conditioned) nonlinear least squares problems from the U.S. National
Institute for Standards and Technology. These may not represent real-world situations.

@@

\texttt{optimx} (John C. Nash and Varadhan (2011)) -- This provides
optimizers that can be applied to minimize a nonlinear function which could be
a nonlinear sum of
squares. NOT generally recommended if nonlinear
least squares programs can be easily used, but provides a check and alternative
solvers.

\subsection{An illustrative example}\label{an-illustrative-example}

The Hobbs weed infestation problem (John C. Nash (1979, 120)) is a growth curve modeling
task that seems straightforward but is quite nasty. Its very succinct statement
provides the ``short reproducible example'' much requested on R mailing lists.
This is a real problem from a field researcher. The data and graph follow.

@@

We thank one of the referees for expressing the viewpoint that a nonlinear regression
approach would use background knowledge and graphs to obtain an understanding of
the general magnitude of the parameters. Indeed, the ``visual fitting'' approach
(John C. Nash and Velleman (1996)) mentioned later accords with this viewpoint, as does use
of the \textbf{Logistic3T} variant of the model.

Our emphasis on the software resilience to starting parameters that are, in the
sense of statistical modeling, ``silly'' comes from over half a century of dealing
with users whose interests and understanding are very far from those of statistical
modelers. Thus we seek methods and codes that obtain reasonable answers under
highly unfavourable conditions. Nevertheless, a proper approach to
nonlinear modeling is to apply all available knowledge to the task.

@@

The usual suggestion is that \(D\) be replaced by a multiple of the unit matrix or else
a multiple of the diagonal part of \(J' J\). In low precision, some elements of
\(J' J\) could underflow to zero (John C. Nash (1977)), and a linear combination of both
choices is an effective compromise. Various choices for
\(D\), as well as a possible line search along the
direction \(\delta\) rather than a unit step (Hartley (1961)),
give rise to several variant
algorithms. ``Marquardt's method'' is a family of methods. Fortunately, most choices
work well.

\subsubsection{Problem setup}\label{problem-setup}

@@

It is also possible to provide R functions for the residual and Jacobian.
This is usually much more work for the user if the formula setup is possible.
To illustrate, we show the functions for the unscaled 3 parameter logistic.
The particular form of these explicit residual and Jacobian functions comes from
their translation from BASIC codes of the 1970s, as adapted in John C. Nash and Walker-Smith (1987). The use
of the Laplace form of the residual and the inclusion of data within the functions
reflects choices of that era that are at odds with current practice. Some users
do still want or need to provide problems as explicit functions, particularly
for problems that are NOT regressions. For example, the Rosenbrock banana-valley
test problem can be provided this way, where the two ``residuals'' are different
functional forms.

\begin{verbatim}
# Logistic3U
hobbs.res  <-  function(x){ # unscaled Hobbs weeds problem -- residual
  if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
  y  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
           38.558, 50.156, 62.948, 75.995, 91.972)
  tt  <-  1:12
  res  <-  x[1]/(1+x[2]*exp(-x[3]*tt)) - y
  # Note: this form of the residual, from Laplace (1788) in the form
  # of "fitted - observed" has been used in our software for half
  # a century, but sometimes offends users of the more traditional
  # "observed - fitted" form
}

hobbs.jac  <-  function(x) { # unscaled Hobbs weeds problem -- Jacobian
  jj  <-  matrix(0.0, 12, 3)
  tt  <-  1:12
  yy  <-  exp(-x[3]*tt)
  zz  <-  1.0/(1+x[2]*yy)
  jj[tt,1]   <-   zz
  jj[tt,2]   <-   -x[1]*zz*zz*yy
  jj[tt,3]   <-   x[1]*zz*zz*yy*x[2]*tt
  attr(jj, "gradient") <- jj
  jj
}
\end{verbatim}

\subsection{Estimation of models specified as formulas}\label{estimation-of-models-specified-as-formulas}

@@

Here we see the infamous ``singular gradient'' termination message of \texttt{nls()}.
Users should, of course, be using the \texttt{SSlogis} \emph{selfStart} tool, but ignorance
of this possibility or a slight variant in the model can easily lead to outcomes
similar to those seen here.

\subsubsection{Solution attempts with nlsr}\label{solution-attempts-with-nlsr}

@@

\subsubsection{Solution attempts with minpack.lm}\label{solution-attempts-with-minpack.lm}

@@

\subsubsection{Solution attempts with gslnls}\label{solution-attempts-with-gslnls}

@@

\subsubsection{Comparison notes for formula-setup solutions}\label{comparison-notes-for-formula-setup-solutions}

@@

\subsection{Functional specification of problems}\label{functional-specification-of-problems}

@@

\subsection{Design goals, termination tests, and output objects}\label{design-goals-termination-tests-and-output-objects}

@@

\subsubsection{Returned results of nls() and other tools}\label{returned-results-of-nls-and-other-tools}

@@

\subsubsection{When to compute ancillary information}\label{when-to-compute-ancillary-information}

Tools that produce a class \texttt{nls} output object create a rich set of functions
and structures that are then used in a variety of modeling tasks, including the
least squares solution. By contrast,
\texttt{nlsr} computes quantities as they are requested or needed, with additional
features in separate functions. For example, the singular values of the Jacobian
are actually computed in the \texttt{print} and \texttt{summary} methods for the result.
These two approaches
lead to different consequences for performance and how features are
provided. \texttt{nlsr} has antecedents in the methods of John C. Nash (1979), where storage for
data and programs was at a ridiculous premium in the small computers of the era.
Thus the code in \texttt{nlsr} is likely of value for workers to copy and modify
for customized tools.

\subsection{Jacobian calculation}\label{jacobian-calculation}

@@

\subsubsection{Jacobian code in selfStart models}\label{jacobian-code-in-selfstart-models}

@@

We are also surprised that the
analytic expressions for the Jacobian (``gradient'') in the \texttt{SSLogis} function
and others save quantities in ``hidden''
variables, i.e., with names preceded by ``.''.
These are then not displayed by the \texttt{ls()}
command, making them difficult to access by users who may wish to create
their own selfStart model via copy and edit.
Interactive tools, such as ``visual fitting'' (John C. Nash and Velleman (1996)) might be
worth considering as another way to find starting parameters, but we know of no
R capability of this type.

@@

\subsection{Bounds constraints on parameters}\label{bounds-constraints-on-parameters}

@@

\subsubsection{Philosophical considerations}\label{philosophical-considerations}

@@

\subsubsection{Fixed parameters (masks)}\label{fixed-parameters-masks}

@@

\subsection{Stabilization of Gauss-Newton computations}\label{stabilization-of-gauss-newton-computations}

@@

\subsubsection{Programming language}\label{programming-language}

@@

\subsection{Data sources for problems}\label{data-sources-for-problems}

@@

\subsubsection{Feature: Subsetting}\label{feature-subsetting}

@@

\subsubsection{Feature: na.action (missing value treatment)}\label{feature-na.action-missing-value-treatment}

@@

\subsection{Feature: model frame}\label{feature-model-frame}

@@

\subsection{Weights on observations}\label{weights-on-observations}

@@

\subsubsection{Weights in returned functions from nls()}\label{weights-in-returned-functions-from-nls}

@@

\subsection{Minor issues with nonlinear least-squares tools}\label{minor-issues-with-nonlinear-least-squares-tools}

\subsubsection{Interim output from the ``port'' algorithm}\label{interim-output-from-the-port-algorithm}

@@

\subsubsection{Failure to return the best result achieved}\label{failure-to-return-the-best-result-achieved}

@@

\subsection{Estimating models that are partially linear}\label{estimating-models-that-are-partially-linear}

@@

\subsection{Models with indexed parameters}\label{models-with-indexed-parameters}

Some models have several common parameters and
others that are tied to particular cases.
The \textbf{man} file for \texttt{nls()} includes an example of a situation in which
parameters are indexed but which uses the ``plinear'' option as an added complication.
Running this example reveals that the answers for the parameters are NOT indexed
as in a vector.
That is, we do not see \texttt{a{[}1{]},\ a{[}2{]},\ a{[}3{]}} but \texttt{a1,\ a2,\ a3}.
This is no doubt because programming for indexed parameters is challenging.
We note that there are capabilities in packages for mixed-effects modeling
such as \texttt{nlme} (Pinheiro et al. (2013)), \texttt{bbmle} (Bolker and Team (2013)), and \texttt{lme4} (D. Bates et al. (2015)) for
estimating models of such type.

\subsection{Tests and use-case examples}\label{tests-and-use-case-examples}

@@

\subsection{Documentation and resources}\label{documentation-and-resources}

@@

\subsection{Future of nonlinear model estimation in R}\label{future-of-nonlinear-model-estimation-in-r}

@@

\section{Acknowledgments}\label{acknowledgments}

@@

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\setlength{\cslentryspacing}{0em}
\begin{CSLReferences}
\bibitem[\citeproctext]{ref-R-onls-manual}
Andrej-Nikolai Spiess. 2022. \emph{Onls: Orthogonal Nonlinear Least-Squares Regression}. \url{https://CRAN.R-project.org/package=onls}.

\bibitem[\citeproctext]{ref-bateswatts}
Bates, D. M., and D. G. Watts. 1988. \emph{Nonlinear Regression Analysis and Its Applications}. Wiley.

\bibitem[\citeproctext]{ref-BatesWatts81}
Bates, Douglas M., and Watts, Donald G. 1981. {``A Relative Offset Orthogonality Convergence Criterion for Nonlinear Least Squares.''} \emph{Technometrics} 23 (2): 179--83.

\bibitem[\citeproctext]{ref-p-lme4}
Bates, Douglas, Martin Mächler, Ben Bolker, and Steve Walker. 2015. {``Fitting Linear Mixed-Effects Models Using {lme4}.''} \emph{Journal of Statistical Software} 67 (1): 1--48. \url{https://doi.org/10.18637/jss.v067.i01}.

\bibitem[\citeproctext]{ref-p-nlstools}
Baty, Florent, and Marie-Laure Delignette-Muller. 2013. \emph{Nlstools: Tools for Nonlinear Regression Diagnostics}.

\bibitem[\citeproctext]{ref-nlsMicrobio-manual}
---------. 2014. \emph{nlsMicrobio: Data Sets and Nonlinear Regression Models Dedicated to Predictive Microbiology}.

\bibitem[\citeproctext]{ref-p-bbmle}
Bolker, Ben, and R Development Core Team. 2013. \emph{Bbmle: Tools for General Maximum Likelihood Estimation}. \url{http://CRAN.R-project.org/package=bbmle}.

\bibitem[\citeproctext]{ref-gslnls23}
Chau, Joris. 2023. \emph{Gslnls: GSL Nonlinear Least-Squares Fitting}. \url{https://CRAN.R-project.org/package=gslnls}.

\bibitem[\citeproctext]{ref-nlshelper-manual}
Duursma, Remko. 2017. \emph{Nlshelper: Convenient Functions for Non-Linear Regression}. \url{https://CRAN.R-project.org/package=nlshelper}.

\bibitem[\citeproctext]{ref-minpacklm12}
Elzhov, Timur V., Katharine M. Mullen, Andrej-Nikolai Spiess, and Ben Bolker. 2012. \emph{Minpack.lm: R Interface to the Levenberg-Marquardt Nonlinear Least-Squares Algorithm Found in MINPACK, Plus Support for Bounds}. R Project for Statistical Computing. \url{http://CRAN.R-project.org/package=minpack.lm}.

\bibitem[\citeproctext]{ref-GSL-manual}
Galassi, Mark, Jim Davies, James Theiler, Brian Gough, and Gerard Jungman. 2009. \emph{GNU Scientific Library - Reference Manual, Third Edition, for GSL Version 1.12 (3. Ed.).}

\bibitem[\citeproctext]{ref-Bioconductor}
Gentleman, Robert C., Vincent J. Carey, Douglas M. Bates, Ben Bolstad, Marcel Dettling, Sandrine Dudoit, Byron Ellis, et al. 2004. {``Bioconductor: Open Software Development for Computational Biology and Bioinformatics.''} \emph{{Genome Biology}} 5 (R80). \url{https://doi.org/10.1186/gb-2004-5-10-r80}.

\bibitem[\citeproctext]{ref-Golub1973}
Golub, G. H., and V. Pereyra. 1973. {``The Differentiation of Pseudo-Inverses and Nonlinear Least Squares Problems Whose Variables Separate.''} \emph{SIAM Journal of Numerical Analysis} 10 (2): 413--32.

\bibitem[\citeproctext]{ref-nls2-22}
Grothendieck, G. 2022. \emph{Nls2: Non-Linear Regression with Brute Force}. \url{https://CRAN.R-project.org/package=nls2}.

\bibitem[\citeproctext]{ref-Hartley1961}
Hartley, H. O. 1961. {``The Modified Gauss-Newton Method for the Fitting of Non-Linear Regression Functions by Least Squares.''} \emph{Technometrics} 3: 269-\/-280.

\bibitem[\citeproctext]{ref-Huet2004}
Huet, S., A. Bouvier, M.-A. Poursat, and E. Jolivet. 2004. \emph{Statistical Tools for Nonlinear Regression: A Practical Guide with {S-PLUS} Examples, 2nd Edition}. Berlin \& New York: Springer-Verlag.

\bibitem[\citeproctext]{ref-nlsr2023manual}
John C Nash, and Duncan Murdoch. 2023. \emph{{nlsr: Functions for Nonlinear Least Squares Solutions}}.

\bibitem[\citeproctext]{ref-tvrdik2007}
Josef Tvrdík and Ivan Křivý and Ladislav Mišík. 2007. {``Adaptive Population-Based Search: Application to Estimation of Nonlinear Regression Parameters.''} \emph{{Computational Statistics \& Data Analysis}} 52 (2): 713--24. \url{https://doi.org/10.1016/j.csda.2006.10.014}.

\bibitem[\citeproctext]{ref-Levenberg1944}
Levenberg, Kenneth. 1944. {``A Method for the Solution of Certain Non-Linear Problems in Least Squares.''} \emph{Quarterly of Applied Mathematics} 2: 164-\/-168.

\bibitem[\citeproctext]{ref-Marquardt1963}
Marquardt, Donald W. 1963. {``{An Algorithm for Least-Squares Estimation of Nonlinear Parameters}.''} \emph{SIAM Journal on Applied Mathematics} 11 (2): 431--41.

\bibitem[\citeproctext]{ref-MiguezNLRAA2021}
Miguez, Fernando. 2021. \emph{{nlraa: Nonlinear Regression for Agricultural Applications}}. \url{https://CRAN.R-project.org/package=nlraa}.

\bibitem[\citeproctext]{ref-jn77ima}
Nash, John C. 1977. {``Minimizing a Nonlinear Sum of Squares Function on a Small Computer.''} \emph{Journal of the Institute for Mathematics and Its Applications} 19: 231--37.

\bibitem[\citeproctext]{ref-cnm79}
---------. 1979. \emph{Compact Numerical Methods for Computers: Linear Algebra and Function Minimisation}. Bristol: Adam Hilger.

\bibitem[\citeproctext]{ref-NashWires22}
---------. 2022. {``{Function minimization and nonlinear least squares in R}.''} \emph{WIREs Computational Statistics}, no. e1580. https://doi.org/\url{https://doi.org/10.1002/wics.1580}.

\bibitem[\citeproctext]{ref-PkgFromRbase22}
Nash, John C., and Arkajyoti Bhattacharjee. 2022. {``Making a Package from Base {R} Files.''} \emph{R-Bloggers}, January. \url{https://www.r-bloggers.com/2022/01/making-a-package-from-base-r-files/}.

\bibitem[\citeproctext]{ref-p-optimx}
Nash, John C, and Ravi Varadhan. 2011. \emph{\emph{Optimx}: A Replacement and Extension of the {optim()} Function}. Nash Information Services Inc.; Johns Hopkins University.

\bibitem[\citeproctext]{ref-nash1996nonlinear}
Nash, John C., and Paul Velleman. 1996. {``Nonlinear Estimation Combining Visual Fitting with Optimization Methods.''} In \emph{Proceedings of the Section on Physical and Engineering Sciences of the American Statistical Association}, 256--61. American Statistical Association.

\bibitem[\citeproctext]{ref-jnmws87}
Nash, John C., and Mary Walker-Smith. 1987. \emph{Nonlinear Parameter Estimation: An Integrated System in BASIC}. Book. New York: Marcel Dekker.

\bibitem[\citeproctext]{ref-NISTnls-manual}
National Institutes for Standards, original from, and Technology http://www.itl.nist.gov/div898/strd/nls/nls\_main.shtml R port by Douglas Bates. 2012. \emph{NISTnls: Nonlinear Least Squares Examples from NIST}. \url{https://CRAN.R-project.org/package=NISTnls}.
